Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-10-14 manuscript.tex
minor edits, pdf

Several demographic factors can modulate or accentuate differences in polymorphism levels maintained on sex chromosomes \citep{ellegren2009}. These include population subdivision and sex-biased dispersal, deviations from a 1:1 breeding sex ratio, and high variance in reproductive success \citep{caballero1995,charlesworth2001,laporte2002,pool2007}. For example, studies comparing levels of variability on the X chromosome and autosomes in humans have found evidence that sex-biased dispersal has reduced levels of X-to-autosome diversity below the neutrally expected level of 0.75 \citep{keinan2009}, though there is considerable variation in estimates of the X/A ratio in humans, and other demographic scenarios have been suggested \citep{hammer2010,bustamante2009}. Theoretical work also indicates that population bottlenecks should lead to a disproportionate decrease in sex-linked compared to autosomal variation \citep{pool2007}, as discovered in chimpanzees and orangutans \citep{kaessmann2001,fischer2006}, the house mouse \textit{Mus musculus} \citep{baines2007}, Drosophila \citep{andolfatto2001}, and humans \citep{keinan2009}.

In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by linked selection, including selective sweeps of benficial mutations \citep{smith1974hitch,aquadro1994} and the removal of strongly deleterious mutations \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}. Consistent with this, studies examining Y chromosome variability in mammals \citep{hellborg2004,bachtrog2013NRG,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have revealed that Y-linked diversity is considerably lower than predicted under models of neutral evolution.

Although... associations among selected alleles in regions of suppressed recombination is expected to lower the coalescent $N_{e}$ for Y-linked genes, so that they behave as if they were subject to weaker selection \citep{hill1966HReffect, mcvean2000}... thereby reducing their effects on linked neutral or nearly neutral variants. This reduction in $N_{e}$ and the effecacy of selection, as a result, reduced levels of neutral polymorphism, selection efficacy, and rate of adaptation \citep{comeron2008}.

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}, studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the HR-effect.

>>>>

Several demographic factors are expected to modulate or accentuate differences in neutral polymorphism maintained on sex chromosomes \citep{ellegren2009}. These include population subdivision and sex-biased dispersal, deviations from a 1:1 breeding sex ratio, and high variance in reproductive success \citep{caballero1995,charlesworth2001,laporte2002,pool2007}. In theory, these processes all result in asymmetries in female vs male $N_{e}$, that could lead to different levels of neutral polymorphism on the sex chromosomes than are expected based on the autosomes (\citep{pool2007}). As evidence for this, coalescent simulations have shown that models of male-biased migration during human out-of-africa dispersal result in levels of X/A neutral diversity simular to what is observed \citep{keinan2009} (though it is worth noting that there is considerable variation in current estimates of X/A in humans, and other demographic scenarios have been suggested \citep{hammer2010,bustamante2009}). In addition, disproportionate losses of sex-linked compared to autosomal variation arising from population bottleneck events have been inferreed in chimpanzees and orangutans \citep{kaessmann2001,fischer2006}, the house mouse \textit{Mus musculus} \citep{baines2007}, and Drosophila \citep{andolfatto2001}, highlighting the importance of demography in affecting patterns of sex-linked and autosomal neutral variation.

In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which refers to the reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. Two types of HRI include (i) selection on stronlgy benficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} and (ii) selection against strongly deleterious mutations
) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}. In both processes, linkage  disequalibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absense of selection.

Consistent with this, studies examining Y chromosome variability in mammals \citep{hellborg2004,bachtrog2013NRG,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have revealed that Y-linked diversity is considerably lower than predicted under models of neutral evolution, with purifying selection playing a key role.

In addition to reducing neutral diversity due to reduced Ne, high linkage disequilibrum (LD) can also occur between the sites experiencing selection such that the net effect that selection has on reducing neutral (or nearly neutral) variation at linked sites is reduced \citep{hill1966HReffect, mcvean2000,comeron2008}. This "weak Hill-Robertson Interference" (weak-HRI) effect is predicted to be greatest in large genomic regions where recombination is rare or absent (\X cite kaiser paper), making Y chromosomes important for empirically investigating this processes.

%discussion point?:
%Moreover, because recombination suppression was relatively recent on plant Y chromosomes, it is an open question whether weak-HRI has had enough time to have a large effect. Although selective interference (indlucing BGS and Selective Sweeps) is expected to have the greatest effects during the early stages of Y-chromosome evolution \citep{bachtrog2008temporal} the R. hastatulus Y chromosome has undergone significant degeneration and gene loss over a short period of time, suggesting that it may behave more similarly to more ancient Y chromosomes.

%Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes , studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the HR-effect.
